---
title: "ESM 244 Lab 7: Working with spatial data"
author:
  - Nathan Grimes
  - Yutian Fang
  - Casey O'Hara
  - Allison Horst
format: 
  html:
    code-fold: show
    toc: true
    number-sections: true
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r}
library(tidyverse)
library(here)
library(broom)

# Spatial data packages
library(sf)
```

# Overview: Spatial data wrangling and visualization

## California county outlines (polygons)

### Read it in with `read_sf`

First, let's read in the California county shapefile:

```{r}
ca_counties_raw_sf <- read_sf(here("data", "ca_counties", "CA_Counties_TIGER2016.shp"))
```

Note we only have to point to the .shp file, not all the others that are related.

## Do a bit of wrangling

Use `View(ca_counties)` to check out what it contains. Let's simplify it by first converting area to km^2^ instead of m^2^, then only keeping county name and area.

All `tidyverse` and `dplyr` functions work the same as before!

```{r}
ca_counties_sf <- ca_counties_raw_sf |> 
```

What if we wanted just the dataframe, without the geometry? Convert to dataframe and select out the geometry column:

```{r}
ca_counties_df <- ca_counties_sf |> 
```

## Check and set the CRS

Use `st_crs()` to check the existing CRS for spatial data. We see that this CRS is "pseudo-mercator" based on WGS 84 - primarily used for web mapping, not analysis. WGS84 (epsg:3857), also note proj4 string and WKT definitions.

```{r}
ca_counties_sf |>  st_crs()
```

## Explore it visually

Plot the California counties using `geom_sf()`. Notice that we can update aesthetics just like we would for a regular ggplot object. Here, we update the color based on land area (and change the color gradient).

# Invasive red sesbania records (spatial points)

## Read in the sesbania data

The data exist in `data/red_sesbania`, and the shapefile is stored as `ds80_for_lab.gpkg`. We also want to check the coordinate reference system, and plot the data for a quick visual understanding.

```{r}
sesbania_sf <- read_sf(here("data","red_sesbania","ds80_for_lab.gpkg")) |> 
  janitor::clean_names()

```

## The analysis:

Let's say we want to find the count of red sesbania observed locations in this dataset *by county*. We then want to create a map of all CA counties using fill color to indicate sesbania counts.

Write pseudocode to accomplish this.

### Pseudocode

## Transform CRS so records match counties

Notice that this CRS is different from the California counties CRS, so we'll want to update it to match. Use `st_transform()` to update the CRS:

```{r}
### if you know the EPSG code:
sesbania_3857_sf <- st_transform(sesbania_sf, 3857)


### if you don't know the EPSG code:
sesbania_3857_2_sf <- st_transform(sesbania_sf, st_crs(ca_counties_sf))

# Then check it: 
sesbania_3857_sf |>  st_crs()
```

Cool, now they have the same CRS.

### Plot them together!

## Spatial join

How can I go about joining these data so that I can find counts by county? Don't worry...`st_join()` has you covered for spatial joins!

```{r}
ca_sesb_sf <- ca_counties_sf |> 
  st_join(sesbania_3857_sf)

head(ca_sesb_sf)
```

We can also join the other direction:

```{r}
sesb_counties_sf <- sesbania_3857_sf |> 
  st_join(ca_counties_sf)

head(sesb_counties_sf)
```

This one keeps the point geometry of the plant records, but appends the county name and area from the CA map.

## Summarize observations by county

Use the CA polygons`sf` object with the CA polygons to summarize the total number of sesbania in each. Watch out for `NAs`!

```{r}
sesb_counts_sf <- ca_sesb_sf |> 
  group_by(county) |> 
  summarize(n_records = XXXXX )

```

Then we can plot a choropleth using the number of records for red sesbania as the fill color (instead of what we used previously, land area):

Make a choropleth map (filled in spatial map) using the number of records for red sesbania as the fill color (instead of what we used previously, land area):

```{r}
ggplot(data = sesb_counts_sf) +
  geom_sf(aes(fill = n_records), color = "white", size = 0.1) +
  scale_fill_gradientn(colors = c("lightgray","orange","red")) +
  theme_minimal() +
  labs(fill = "Number of S. punicea records")
```

Now, on your own let's practice our spatial data wrangling!

Only plot the county with the greatest number of red sesbania records, and make a map of those locations.

Hint: It's okay to break the steps apart, you don't need to do them all in one step (in fact it's probably preferred you don't).

### Pseudocode here!
